\documentclass{book}
\input{ASLmacros}
\usepackage{changebar}

\newcommand\lrmexcerpt[2]{\cbstart #1 \cbend}

\author{Arm Architecture Technology Group}
\title{ASL Programmer's Guide \\
       DDI ????}
\begin{document}
\maketitle

\tableofcontents{}

\include{notice.tex}

\include{disclaimer.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Preface}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
List the reference documents (Syntax Reference, Typing Reference, Semantics Reference).

Reference the LRM.

\section{Conventions}

\section{Language Versioning}
Language versioning stuff?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lrmexcerpt{
The \emph{Architecture Specification Language} (ASL) is designed and used to specify architectures. As a formal
specification language it is designed to be accessible, understandable and unambiguous to programmers, hardware
engineers and hardware verification engineers who collectively have quite a small intersection of languages they
all understand. It can intentionally under specify behaviors in the architecture being described.

ASL is:
\begin{itemize}
\item a first-order language with strong static type-checking.
\item whitespace-insensitive.
\item imperative.
\end{itemize}

ASL has support for:
\begin{itemize}
\item bitvectors:
  \begin{itemize}
    \item as a type.
    \item as a literal constant.
    \item bitvector concatenation.
    \item bitvector constants with wildcards.
    \item bitslices.
    \item dependent types to support function overloading using bitvector lengths.
    \item dependent types to reason about lengths of bitvectors.
  \end{itemize}
\item unbounded arithmetic types ``integer'' and ``real''.
\item type inference.
\item exceptions.
\item enumerations.
\item arrays.
\item records.
\end{itemize}

ASL does not have support for:
\begin{itemize}
    \item references or pointers.
\end{itemize}

A \emph{specification} consists of a self-contained collection of ASL code.
}{16--17}

Example specifications...

\section{Available Tools}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Data Types}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Integers}

\section{Reals}

\section{Booleans}

\section{Strings}

\section{Enumerations}

\section{Bitvectors}

\section{Arrays}

\section{Tuples}

\section{Records}

\section{Exceptions}

\section{Subtyping}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Variable Declarations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Variables can be:
\begin{itemize}
    \item Constant
    \item Mutable
    \item Immutable
    \item Configuration
    \item Local
    \item Global
    \item Initialized/uninitialized
    \item Primitive
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Expressions}

\section{Operators}

\section{Record Expressions}

\section{Slicing}

\section{Tuples}

\section{Conditionals}

\section{Pattern Matching}

\section{Checked Type Conversions}

\section{UNKNOWN}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Statements}

\section{Assignments}

\section{Conditionals}

\section{Cases}

\section{Loopings}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Subprograms}

Subprograms can be:
\begin{itemize}
    \item Functions
    \item Procedures
    \item Getters
    \item Setters
    \item Recursive with a recursion limit annotation
    \item Primitive
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Exception Handling}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Standard Library}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Primitives}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Configurations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Concurrency}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\bibliography{ASL}
\end{document}