\chapter*{Changelog}

\section*{EAC2}

The following changes were introduced in version EAC2.

\subsection*{AARCH-25136: Improve exposition of ASL1.0's grammar construction}

The construction of ASL1.0's grammar has been clarified in the following sections:
\begin{itemize}
  \item Introduction to \chapref{Syntax}
  \item \secref{ASLGrammar}
  \item \secref{PriorityAndAssociativity}
  \item \secref{GrammarExamples}, a new section with worked examples
  \item \secref{BinaryExpressions}
\end{itemize}

\subsection*{AARCH-25137: Improve exposition of ASL1.0's evaluation order and its consequences}

Order of evaluation has been clarified in the following sections:
\begin{itemize}
  \item \secref{EvaluationOrder}, including new example \ExampleRef{Evaluation order}
  \item \secref{ParenthesisedExpressions}
\end{itemize}

\subsection*{AARCH-25350: Remove ASL1.0 shorthands for readability}

The following shorthand slice forms are no longer permitted:
\begin{itemize}
  \item Scaled form, \lstinline{factor*:len}. Equivalent to \lstinline{(factor*len) +: len}.
  \item Shorthand length form, \lstinline{:len}. Equivalent to \lstinline{0 +: len}.
\end{itemize}

Eliding of parameters on the right-hand sides of declarations with an explicit \lstinline{bits(N)} type annotation is no longer permitted.
For example, the following is no longer valid ASL:
\begin{lstlisting}
let x : bits(32) = Zeros{}();
let y : bits(64) = ZeroExtend{,1}('1');
\end{lstlisting}

\subsection*{AARCH-25640: Remove ASL1.0 constructs not used in pseudocode}

The following features are removed:
\begin{itemize}
  \item Enumeration-indexed arrays.
  \item Subtyping.
  \item Nested patterns.
\end{itemize}

\paragraph{Enumeration-indexed-arrays.}

Arrays of the form
\begin{lstlisting}
array [[enumeration_type]] of element_type
\end{lstlisting}
are no longer permitted.
For example, the following is no longer valid ASL:
\begin{lstlisting}
type Color of enumeration {RED, GREEN, BLUE};
var x : array[[Color]] of integer;
\end{lstlisting}

\paragraph{Subtyping.}

Named type declarations using the \lstinline{subtypes} or \lstinline{with} keywords are no longer permitted.
For example, the following is no longer valid ASL:
\begin{lstlisting}
type SuperInt of integer;
type SubInt of integer subtypes SuperInt;

type Packet of record { data : bits(32) };
type TimedPacket subtypes Packet with { time : integer };
\end{lstlisting}

\paragraph{Nested patterns.}

ASL patterns of the following set-like forms are no longer permitted to be nested:
\begin{lstlisting}
{pattern1, ...}
!{pattern1, ...}
\end{lstlisting}
ASL nested patterns of the following tuple form are no longer permitted:
\begin{lstlisting}
(pattern1, ...)
\end{lstlisting}
For example, the following will no longer be valid ASL.
\begin{lstlisting}
x IN {{1, 2}, 3}
y IN {!{TRUE}}
z IN {(1, 2)}
\end{lstlisting}
Note that this does not affect tuple expressions of the form:
\begin{lstlisting}
(expression1, ...)
\end{lstlisting}
