%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Basic Type Attributes\label{sec:BasicTypeAttributes}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section defines some basic predicates for classifying types as well as
functions that inspect the structure of types:
% TYPING_RULES_BEGIN
% This is generated code; run `python3 doclint.py -cm` to update.
\begin{itemize}
  \item \TypingRuleRef{BuiltinSingularType} Builtin singular types.
  \item \TypingRuleRef{NamedType} Named types.
  \item \TypingRuleRef{AnonymousType} Anonymous types.
  \item \TypingRuleRef{SingularType} Singular types.
  \item \TypingRuleRef{StructuredType} Structured types.
  \item \TypingRuleRef{Structure} The structure of a type.
  \item \TypingRuleRef{MakeAnonymous} The underlying type of a type.
  \item \TypingRuleRef{CheckConstrainedInteger} Checked constrained integers.
\end{itemize}
% TYPING_RULES_END

\TypingRuleDef{BuiltinSingularType}
\RuleDescription{Builtin singular types.}
\RenderRelation{is_builtin_singular}

\ExampleDef{Builtin singular types}
\listingref{typing-builtinsingulartype} defines variables of builtin singular types
\texttt{integer}, \texttt{real},
\texttt{boolean}, \texttt{bits(4)}, and~\texttt{bits(2)}
\ASLListing{Examples of builtin singular types}{typing-builtinsingulartype}{\typingtests/TypingRule.BuiltinSingularTypes.asl}

\ExampleDef{Builtin enumeration types}
In \listingref{typing-builtinenumerationtype},
the builtin singular type \texttt{Color} consists of two constants:
\texttt{RED} and~\texttt{BLACK}.
\ASLListing{An enumeration type}{typing-builtinenumerationtype}{\typingtests/TypingRule.EnumerationType.asl}

\RenderProseAndFormally{is_builtin_singular}
\CodeSubsection{\BuiltinSingularBegin}{\BuiltinSingularEnd}{../types.ml}


\TypingRuleDef{NamedType}
\RuleDescription{Named types.}
\RenderRelation{is_named}

\Enumerationtypesterm{}, record types, collection types, and exception types
must be declared and associated with a named type.

\ExampleDef{Named Types}
In the specification

In \listingref{named-types2},
\texttt{qualifiedData} and \texttt{DatawithFlag} both define named types.

\RenderProseAndFormally{is_named}
\CodeSubsection{\NamedBegin}{\NamedEnd}{../types.ml}


\TypingRuleDef{AnonymousType}
\RuleDescription{Anonymous types.}
 %
\RenderRelation{is_anonymous}

\ExampleDef{Anonymous Types}
The well-typed specification in \listingref{AnonymousType} illustrates the use
of anonymous types as permitted by \TypingRuleRef{TypeSatisfaction}.
\ASLListing{Well-typed anonymous types}{AnonymousType}{\typingtests/TypingRule.AnonymousType.asl}

\RenderProseAndFormally{is_anonymous}
\CodeSubsection{\AnonymousBegin}{\AnonymousEnd}{../types.ml}

\TypingRuleDef{SingularType}
\RuleDescription{Singular types.}
%
\RenderRelation{is_singular}

\ExampleDef{Singular types}
In \listingref{SingularType}, the types \texttt{A}, \texttt{B}, and \texttt{C} are all singular types:

\ASLListing{Singular types}{SingularType}{\typingtests/TypingRule.SingularType.asl}

\RenderProseAndFormally{is_singular}
\CodeSubsection{\SingularBegin}{\SingularEnd}{../types.ml}

\TypingRuleDef{StructuredType}
\RuleDescription{Structured types.}
\hypertarget{def-structuredtype}{}
A \emph{\structuredtypeterm} is any type that consists of a list of field
identifiers that denote individual storage elements.
In ASL there are three such types --- record types, collection types, and
exception types.

\RenderRelation{is_structured}

\ExampleDef{Structured Types}
In \listingref{StructuredTypes}, the types \texttt{SyntaxException} and \texttt{PointRecord}
are each an example of a \structuredtypeterm{}.
\ASLListing{Structure types}{StructuredTypes}{\typingtests/TypingRule.StructuredType.asl}

\RenderProseAndFormally{is_structured}

\TypingRuleDef{Structure}
\RuleDescription{The structure of a type.}
\RenderRelation{get_structure}

\TypingRuleRef{TypeCheckAST} ensures the absence of circular type definitions,
which ensures that \TypingRuleRef{Structure} terminates\footnote{In mathematical terms,
this ensures that \TypingRuleRef{Structure} is a proper \emph{structural induction.}}.

\ExampleDef{The Structure of a Type}
\listingref{Structure} shows examples of types and their structures.
\ASLListing{Types and their structure}{Structure}{\typingtests/TypingRule.Structure.asl}


\RenderProseAndFormally{get_structure}
\CodeSubsection{\StructureBegin}{\StructureEnd}{../types.ml}


\TypingRuleDef{MakeAnonymous}
\RuleDescription{The underlying type of a type.}
\RenderRelation{make_anonymous}

\ExampleDef{The Underlying Type of a Type}
\ASLListing{Anonymous types}{MakeAnonymous}{\typingtests/TypingRule.MakeAnonymous.asl}

In \listingref{MakeAnonymous}, the \underlyingtypesterm\ of \texttt{integer}, \texttt{T1}, and \texttt{T2} is \texttt{integer}.
Further, the \underlyingtypeterm{} of \texttt{(integer, T2)} and \texttt{T3} is
\texttt{(integer, T2)}.  Notice how the \underlyingtypeterm{} does not replace
\texttt{T2} with its own \underlyingtypeterm, in contrast to the \structureterm{} of
\texttt{T2}, which is \texttt{(integer, integer)}.

\RenderProseAndFormally{make_anonymous}
\CodeSubsection{\MakeAnonymousBegin}{\MakeAnonymousEnd}{../types.ml}

\TypingRuleDef{CheckConstrainedInteger}
\RuleDescription{Checked constrained integers.}
A type is a \emph{\constrainedintegerterm} if it is either a \wellconstrainedintegertypeterm{}
or a \parameterizedintegertypeterm.

\RenderRelation{check_constrained_integer}

\ExampleDef{Checking for Constrained Integers}
\listingref{check-constrained-integer}
shows examples of checking whether a type (used for the width of a bitvector type)
is a \constrainedintegerterm{} type.
\ASLListing{Checking for constrained integers}{check-constrained-integer}{\typingtests/TypingRule.CheckConstrainedInteger.asl}


\RenderProseAndFormally{check_constrained_integer}
\CodeSubsection{\CheckConstrainedIntegerBegin}{\CheckConstrainedIntegerEnd}{../Typing.ml}
